---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---


## Loading and preprocessing the data

```{r message=FALSE}
library(chron)
library(lubridate)
library(dplyr)
library(ggplot2)
library(xtable)
```

## The section below deals with reading in the csv file, creating a new data-time variable. Changing the formats of 
## interval and Date into time and Date class.

```{r}
filepath<- "E:\\johns hopkins\\RepData_PeerAssessment1-master\\activity.csv"
f1<- read.csv(file =filepath,stringsAsFactors = FALSE )

# convert date from vector of charachters into date object.
f1$date<- as.Date(x = f1$date,format ="%Y-%m-%d")
# interval in to be properly formatted in the TIME class instead of being a integer.
f1$interval<-sprintf(fmt = "%04d",f1$interval)
# ensuring 04 above, will left pad with zeros.
# now use the format function.
f1$interval<-format(x =strptime(x = f1$interval,format = "%H%M"),format = "%H:%M")
# the strptime --> gives a POSIXlt object with the current time. This date portion is striped out by the format command.
f1$interval<-paste(f1$interval,":00",sep = "")
# convert the interval into a time object(chron)
f1$interval<-times(x = f1$interval)

# create a new variable datetime. This will then be useful for plotting on the x-axis.
# combine the date and time into a single new column
f1$date_time<-ymd_hms(paste(f1$date,f1$interval))

```


## What is mean total number of steps taken per day?
### The data is taken over a two month period. We shall aggregate on a per day basis.
```{r}


 
# to find the mean/median of the total number of steps /day.
rsum<-aggregate(formula=f1$steps~f1$date,data = f1,FUN = sum)
#rsum
# WE need to calulate the mean of the sum( total number of steps /day)
mean_total_steps_day<- mean(x = rsum$`f1$steps`,na.rm = TRUE)
#mean_total_steps_day
# similarly for median
median_total_steps_day<-median(x = rsum$`f1$steps`,na.rm = TRUE)
# construct the histogram
 g<- ggplot(data = rsum,mapping = aes(x = rsum$`f1$steps`))
 g+geom_histogram()+stat_bin(binwidth = 500)+xlab("Total number of steps per day")

```

The  mean total number of steps per day is `r mean_total_steps_day` .<br>
The median total number of steps per day is `r median_total_steps_day`.<br>


## What is the average daily activity pattern?

```{r}
# make a plot
g<- ggplot(data = f1,mapping = aes(x = f1$date_time,y = f1$steps))
g+geom_line()
```

```{r}
result<-f1%>%filter(f1$steps==max(f1$steps,na.rm = TRUE))
#print(result)
max_steps <-result$steps
rinterval<-result$date_time

```
The max number of steps are `r max_steps`. The 5 minute interval is `r rinterval`

## Imputing missing values

```{r}
 rcount<-sum(is.na(f1$steps))
```

## The total number of missing values equals `r rcount`.

### We shall fill the missing data using the median value, since the distribution is skewed.
```{r}
# insert the median value.
f2<- f1%>%group_by(f1$date)%>%summarise(Median=median(f1$steps,na.rm = TRUE))
#print(f2)
f1$steps<-replace(f1$steps,is.na(f1$steps),values = f2$Median)

```
We shall now find the summary stats with the new imputed values.
```{r}



rsum1<-aggregate(formula=f1$steps~f1$date,data = f1,FUN = sum)
#rsum1
# WE need to calulate the mean of the sum( total number of steps /day)
mean_total_steps_day1<- mean(x = rsum1$`f1$steps`,na.rm = TRUE)
#mean_total_steps_day
# similarly for median
median_total_steps_day1<-median(x = rsum1$`f1$steps`,na.rm = TRUE)

# construct the histogram
 g<- ggplot(data = rsum1,mapping = aes(x = rsum1$`f1$steps`))
 g+geom_histogram()+stat_bin(binwidth = 500)+xlab("Total number of steps per day")

```









We have imputed the median value to replace the NAs.The median in most cases was 0.

<br>The new mean total number of steps/day and new median total number of steps /day areas follows:
<br>mean: `r mean_total_steps_day1`
<br>median:`r median_total_steps_day1`

## Are there differences in activity patterns between weekdays and weekends?

```{r}
# create a new factor variable.
f1<-mutate(f1,type_day =ifelse(weekdays(x = f1$date,abbreviate = FALSE) %in% c('Saturday','Sunday'),yes = 'Weekend',no = 'weekday'))
f1$type_day<-factor(f1$type_day)
```

We shall make a time-series plot using the ggplot2 system
```{r}
g<- ggplot(data = f1,mapping = aes(x = f1$date_time,y = f1$steps))
g+geom_line()+ facet_grid(f1$type_day~.)

f1%>%group_by(f1$type_day)%>%summarize(Average_no_of_steps = mean(f1$steps,na.rm = TRUE),standard_deviation = sd(f1$steps,na.rm = TRUE))
```

As shown by the summary table above there is no difference in the average activity on a weekday as compared to a weekend.










